---
output: pdf_document
title: "Modeling Canada Housing Prices"
author: "Cole Chisamore"
---

```{r, echo=FALSE}
setwd("C:/Users/colef/Documents/GitHub/Housing-Web-Scraping")
```

# Introduction

In this pdf, I will using R to attempt to model the response variable housing price based on multiple explanatory variables:

* Bedrooms: Number of bedrooms in the property (main or otherwise)
* Bathrooms: Number of bathrooms in the property (main or otherwise)
* Square Feet: Rough square footage of indoor property
* Property Type: Category assigned to each listing including...
  - Apt/Condo
  - Duplex
  - Fourplex
  - House
  - Land/Lot
  - Mfd/Mobile Home
  - Multifamily
  - Recreational
  - Townhouse
  - Triplex


# Forming the Initial Model

Let's use some libraries first:

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(ggsci)
```

Then we can import the raw scraped housing data that was collected beforehand:

```{r}
# Read in csv file as a tibble,
houses.df <- read_csv(file = "housing_df.csv", show_col_types = FALSE)
# remove any data with invalid values,
houses.df <- houses.df %>%
  drop_na()
# get rid of duplicates (some listings appeared multiple times),
houses.df <- houses.df %>%
  distinct()
# make Property Type a factor, and order it
houses.df$`Property Type` <- factor(houses.df$`Property Type`)
houses.df <- houses.df %>%
  mutate(`Property Type` = fct_reorder(`Property Type`, Price, 
                                       .fun="length", .desc = TRUE))
```

Let's see an initial naive model:

```{r}
houses.glm1 <- glm(formula = Price~Bedrooms+Bathrooms+`Square Feet`+`Property Type`,
                   data=houses.df)
summary(houses.glm1)
```

Remarks on initial model:

* the Bedrooms variable is oddly assigned a negative (and significant) correlation with Price, which is counter-intuitive
* multiple Property Type categories are deemed to have insignificant difference from "House"





\newpage

# Refining the Dataframe

Before creating another model, I will be going through each variable (response or explanatory) to refine our dataframe. This includes graphing their distributions and inspecting specific qualities to remove outlying/incorrect data and develop an deeper understanding about our dataset.

## Checking Price

Let's check the distribution of house listing prices, to see if we can remove any outliers.

```{r, fig.width=6, fig.height=3, warning=FALSE}
#' Creates quantiles of middle 99% range of data and distribution plot
#' 
#' @param df A Dataframe
#' @param column A string representing a column name
#' @param datatype A string denoting plot type
#' @return list of quantile values
#' @return distribution plot
distribution_graph <- function(df, column, datatype="bar") {
  q <- quantile(df[[column]], probs=c(0,0.005,0.995,1))
  y <- if(datatype=="density"){geom_density()}else{geom_bar()}
  p <- ggplot(data=df, aes(x=.data[[column]])) + y +
    ggtitle(paste("Distribution of", column)) +
    geom_vline(xintercept=q[c(2,3)], linetype="dashed", color="red") + 
    geom_vline(xintercept=q[c(1,4)], linetype="dashed", color="blue")
  return(list(quantiles=q,plot=p))
}
distribution_graph(houses.df, "Price","density")
```

This distribution is a little loo skewed by some extremely expensive house listings. Since 99% of the price values lie within the prices of $80,000 to $5,000,000, let's remove the values which go beyond this range to make the graph more normally distributed.

\newpage 

```{r, fig.width=6, fig.height=2.8, warning=FALSE}
# filter out outlying prices and redo plotting:
houses.df <- houses.df %>%
  filter(Price <= 5000000) %>%
  filter(Price >= 80000)
price.plot <- distribution_graph(houses.df, "Price","density")
price.plot$quantiles
price.plot$plot
```

This distribution is now a lot more normal for our model to handle. However the right-skewedness suggests that a transformation on Price would be beneficial to our model.

\newpage

## Checking Square Feet

Let's continue by checking the distribution of house square footage.
```{r, fig.width=6, fig.height=2, warning=FALSE}
distribution_graph(houses.df, "Square Feet","density")
```

Since this distribution already seems relatively normal, I will only remove listings above 10,000 square feet.

```{r, fig.width=6, fig.height=2, warning=FALSE}
# filter out outlying Square Feet and redo plotting:
houses.df <- houses.df %>%
  filter(`Square Feet` <= 10000)
distribution_graph(houses.df, "Square Feet","density")
```

\newpage

## Checking Bathrooms

Now moving onto bathrooms, which seem normal already:
```{r, fig.width=6, fig.height=2, warning=FALSE}
distribution_graph(houses.df, "Bathrooms")
```


## Checking Bedrooms

Finally bedrooms, which also seems normal already:
```{r, fig.width=6, fig.height=2, warning=FALSE}
distribution_graph(houses.df, "Bedrooms")
```

\newpage

## Checking Property Type

Now let's look at the distributions of the categorical variable property types:
```{r}
houses.df %>%
  group_by(`Property Type`) %>%
  summarise(count=n())
ggplot(data=houses.df[houses.df$Price<=3000000,], 
       aes(x=Price, fill=`Property Type`, color=`Property Type`)) +
  geom_density(linewidth=1,alpha=0.1) + ggtitle("Price Boxplot for Each Property Type") +
  theme(legend.position="inside", legend.position.inside = c(0.8, 0.6)) +
  scale_color_d3() + scale_fill_d3()
```
*Note: x-axis has been shrunk to make distributions clearer

\newpage 

Due to small sizes of some categories, I will be merging some categories into one based on similar distribution/property type. First I will merge "Multifamily", "Triplex", and "Fourplex", since each has a wider further-right distribution, and are all properties with multiple living units. Since "Recreational" and "Land/Lot" both have so few occurences combined with unique distribution and property type, I will be removing these listings from our dataset:

```{r}
# multi-family merging
houses.df[houses.df$`Property Type` %in% 
            c("Triplex","Fourplex"),]$`Property Type` <- 
  "Multifamily"

# removing categories
houses.df <- houses.df %>%
  filter(!(`Property Type` %in% c("Land/Lot","Recreational")))
```

Now let's recheck the distributions:
```{r}
ggplot(data=houses.df[houses.df$Price<=3000000,], 
       aes(x=Price, fill=`Property Type`, color=`Property Type`)) +
  geom_density(linewidth=1,alpha=0.1) + ggtitle("Price Boxplot for Each Property Type") +
  theme(legend.position="inside", legend.position.inside = c(0.8, 0.6)) +
  scale_color_d3() + scale_fill_d3()
```

\newpage

# Remodeling

After filtering our dataset and implementing changes to the naive model we arrive at a our new and improved model:

```{r}
houses.glm2 <- glm(formula = Price~Bedrooms+Bathrooms+`Square Feet`+`Property Type`, 
                   data=houses.df, family = Gamma(link="log"))
summary(houses.glm2)
```




